---
title: "Market Dynamics Explorer — Under the Hood"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

This document mirrors every computation the Shiny app performs, but runs
them statically so you can inspect the data and charts step by step.

---

## 1. Setup

```{r}
library(RTL)
library(tidyquant)
library(dplyr)
library(tidyr)
library(lubridate)
library(slider)
library(plotly)
library(broom)
```

---

## 1.1 Feather File Check

```{r}
test <- arrow::read_feather(here::here("inst/extdata/futures_data.feather"))
tes1 <- arrow::read_feather((here::here("inst/extdata/cmt_data.feather")))

# All unique tickers in RTL::dflong
unique(sub("[0-9]+$", "", RTL::dflong$series))
```

---

## 2. Data Loading

### 2.1 RTL Futures Data (`fct_load_futures_data`)

`RTL::dflong` stores continuous futures contracts in long format.
The `series` column encodes market + contract number as a single string
(e.g. `"CL01"` = WTI crude, front month).  We split it into two columns.

```{r}
futures_raw <- RTL::dflong

# Parse "TICKERNN" into separate market and contract columns
# sub("[0-9]+$", ...) strips trailing digits  → market code
# sub("^[A-Za-z]+", ...) strips leading letters → contract number
futures <- futures_raw |>
  mutate(
    market   = sub("[0-9]+$", "", series),
    contract = as.integer(sub("^[A-Za-z]+", "", series)),
    date     = as.Date(date)
  ) |>
  filter(!is.na(value), !is.na(contract)) |>
  select(date, market, contract, value)

# Confirm what we have
cat("Date range:", format(min(futures$date)), "→", format(max(futures$date)), "\n")
cat("Markets:   ", paste(unique(futures$market), collapse = ", "), "\n")
cat("Rows:      ", nrow(futures), "\n")
```

```{r}
# Quick look at one market
futures |>
  filter(market == "CL", date == max(date)) |>
  arrange(contract)
```

### 2.2 FRED Constant Maturity Treasury Data (`fct_load_cmt_data`)

We pull 11 maturities (1 month → 30 years) from FRED.
`tidyquant::tq_get()` does not require an API key for FRED economic data.

```{r}
cmt_meta <- tibble(
  ticker         = c("DGS1MO","DGS3MO","DGS6MO",
                     "DGS1","DGS2","DGS3","DGS5",
                     "DGS7","DGS10","DGS20","DGS30"),
  maturity_years = c(1/12, 3/12, 6/12, 1, 2, 3, 5, 7, 10, 20, 30)
)

cmt <- lapply(seq_len(nrow(cmt_meta)), function(i) {
  tq_get(cmt_meta$ticker[i], get = "economic.data") |>
    rename(value = price) |>
    mutate(
      maturity_label = cmt_meta$ticker[i],
      maturity_years = cmt_meta$maturity_years[i],
      date           = as.Date(date)
    ) |>
    select(date, maturity_label, maturity_years, value)
}) |>
  bind_rows() |>
  filter(!is.na(value))

cat("CMT date range:", format(min(cmt$date)), "→", format(max(cmt$date)), "\n")
cat("Maturities:    ", paste(unique(cmt$maturity_label), collapse = ", "), "\n")
```

---

## 3. Forward Curve

The forward curve shows the **term structure of prices** at a point in time:
how does the market price contracts expiring 1 month, 2 months, ... N months
out?  Backwardation (downward slope) signals tight near-term supply.
Contango (upward slope) reflects storage costs and expectations of future
surplus.

### 3.1 Commodity Forward Curves — Multi-date Overlay

We pick three snapshot dates spaced one year apart to show how the curve
shape has evolved.

```{r}
snapshot_dates <- as.Date(c("2019-01-02", "2022-01-03", "2025-01-02"))
markets_to_show <- c("CL", "BRN", "NG")

# For each market × date, find the closest available trading date ≤ snapshot
get_curve <- function(data, mkt, snap_date) {
  data |>
    filter(market == mkt, date <= snap_date) |>
    filter(date == max(date)) |>
    arrange(contract) |>
    mutate(snapshot = snap_date)
}

curve_df <- bind_rows(lapply(markets_to_show, function(mkt) {
  bind_rows(lapply(snapshot_dates, function(d) get_curve(futures, mkt, d)))
}))

# Opacity ramp: oldest = most transparent
opacities <- c(0.35, 0.65, 1.0)

p_curve <- plot_ly()
for (mkt in markets_to_show) {
  for (i in seq_along(snapshot_dates)) {
    sub_df <- filter(curve_df, market == mkt, snapshot == snapshot_dates[i])
    if (nrow(sub_df) == 0) next
    p_curve <- add_trace(
      p_curve,
      data          = sub_df,
      x             = ~contract,
      y             = ~value,
      type          = "scatter",
      mode          = "lines+markers",
      opacity       = opacities[i],
      name          = paste0(mkt, " (", format(snapshot_dates[i], "%Y-%m-%d"), ")"),
      hovertemplate = paste0(mkt, " C%{x}<br>$%{y:.2f}<extra></extra>")
    )
  }
}
p_curve |>
  layout(
    title     = "Energy Futures Forward Curves",
    xaxis     = list(title = "Contract #"),
    yaxis     = list(title = "Price (USD)"),
    hovermode = "x unified",
    legend    = list(orientation = "h", y = -0.25)
  )
```

### 3.2 US Treasury Yield Curve — Multi-date Overlay

Same idea applied to interest rates.  A flat or inverted yield curve
(short rates ≥ long rates) is a classic recession signal.

```{r}
get_yield_curve <- function(snap_date) {
  cmt |>
    filter(date <= snap_date) |>
    filter(date == max(date)) |>
    arrange(maturity_years) |>
    mutate(snapshot = snap_date)
}

yc_df <- bind_rows(lapply(snapshot_dates, get_yield_curve))

p_yc <- plot_ly()
for (i in seq_along(snapshot_dates)) {
  sub_df <- filter(yc_df, snapshot == snapshot_dates[i])
  p_yc <- add_trace(
    p_yc,
    data          = sub_df,
    x             = ~maturity_years,
    y             = ~value,
    type          = "scatter",
    mode          = "lines+markers",
    opacity       = opacities[i],
    name          = format(snapshot_dates[i], "%Y-%m-%d"),
    hovertemplate = "%{x:.1f}y: %{y:.2f}%<extra></extra>"
  )
}
p_yc |>
  layout(
    title     = "US Constant Maturity Treasury Yield Curves",
    xaxis     = list(title = "Maturity (years)"),
    yaxis     = list(title = "Yield (%)"),
    hovermode = "x unified",
    legend    = list(orientation = "h", y = -0.25)
  )
```

---

## 4. Volatility

### 4.1 Rolling Realised Volatility — Front Month

We compute log returns then apply a rolling standard deviation, annualised
by multiplying by √252 (trading days per year).  High-vol regimes correspond
to geopolitical events, supply shocks, and demand crises.

```{r}
# Rolling annualised vol of front-month (contract = 1) log returns
window_days <- 21  # ~1 trading month

front_vol <- futures |>
  filter(contract == 1) |>
  arrange(market, date) |>
  group_by(market) |>
  mutate(
    log_ret  = c(NA_real_, diff(log(value))),
    roll_vol = slide_dbl(
      log_ret,
      ~ sd(.x, na.rm = TRUE) * sqrt(252),
      .before   = window_days - 1L,
      .complete = TRUE
    )
  ) |>
  ungroup() |>
  filter(!is.na(roll_vol))

p_vol <- plot_ly()
for (mkt in unique(front_vol$market)) {
  sub_df <- filter(front_vol, market == mkt)
  p_vol <- add_trace(
    p_vol,
    data          = sub_df,
    x             = ~date,
    y             = ~roll_vol,
    type          = "scatter",
    mode          = "lines",
    name          = mkt,
    hovertemplate = "%{x|%Y-%m-%d}<br>Vol: %{y:.1%}<extra></extra>"
  )
}
p_vol |>
  layout(
    title     = paste0(window_days, "-Day Rolling Annualised Volatility (Front Month)"),
    xaxis     = list(title = "Date"),
    yaxis     = list(title = "Annualised Vol", tickformat = ".0%"),
    hovermode = "x unified",
    legend    = list(orientation = "h", y = -0.25)
  )
```

### 4.2 Volatility Surface — CL Term Structure

For a single market we compute vol for every contract, then display it as a
heatmap: time on x, contract number on y, colour = vol level.  This reveals
whether near-term or far-term contracts are more volatile (usually near-term),
and whether vol is elevated across the whole curve or concentrated.

```{r}
surface_market <- "CL"

cl_surface <- futures |>
  filter(market == surface_market) |>
  arrange(contract, date) |>
  group_by(contract) |>
  mutate(
    log_ret  = c(NA_real_, diff(log(value))),
    roll_vol = slide_dbl(
      log_ret,
      ~ sd(.x, na.rm = TRUE) * sqrt(252),
      .before   = window_days - 1L,
      .complete = TRUE
    )
  ) |>
  ungroup() |>
  filter(!is.na(roll_vol)) |>
  # Keep Mondays only to reduce matrix size for display
  filter(wday(date) == 2L)

wide_surface <- cl_surface |>
  pivot_wider(id_cols = date, names_from = contract, values_from = roll_vol)

mat       <- as.matrix(wide_surface[, -1])
dates_vec <- wide_surface$date
contracts_vec <- as.integer(colnames(wide_surface)[-1])

plot_ly(
  x             = dates_vec,
  y             = contracts_vec,
  z             = t(mat),
  type          = "heatmap",
  colorscale    = "RdYlGn",
  reversescale  = TRUE,
  hovertemplate = "Date: %{x|%Y-%m-%d}<br>Contract: %{y}<br>Vol: %{z:.1%}<extra></extra>"
) |>
  layout(
    title = paste0(surface_market, " Volatility Surface (", window_days, "-day rolling)"),
    xaxis = list(title = "Date"),
    yaxis = list(title = "Contract #")
  )
```

---

## 5. Co-Dynamics

### 5.1 Correlation Matrix Snapshot

Correlation is computed on log returns of front-month prices.
A high WTI/Brent correlation is expected; understanding when it breaks down
(e.g. US export ban, regional supply disruptions) tells you about
market segmentation.

```{r}
corr_window   <- 63   # ~1 quarter of trading days
corr_snap     <- as.Date("2025-01-02")
corr_markets  <- c("CL", "BRN", "NG", "HO", "RB")

# Wide returns matrix
wide_ret <- futures |>
  filter(market %in% corr_markets, contract == 1) |>
  arrange(market, date) |>
  group_by(market) |>
  mutate(log_ret = c(NA_real_, diff(log(value)))) |>
  ungroup() |>
  filter(!is.na(log_ret)) |>
  pivot_wider(id_cols = date, names_from = market, values_from = log_ret) |>
  arrange(date)

# Slice the window ending at the snapshot date
window_data <- wide_ret |>
  filter(date <= corr_snap) |>
  tail(corr_window)

corr_mat <- cor(window_data[, -1], use = "pairwise.complete.obs")

plot_ly(
  x             = colnames(corr_mat),
  y             = colnames(corr_mat),
  z             = corr_mat,
  type          = "heatmap",
  colorscale    = "RdBu",
  reversescale  = TRUE,
  zmid = 0, zmin = -1, zmax = 1,
  hovertemplate = "%{x} vs %{y}<br>Corr: %{z:.2f}<extra></extra>"
) |>
  layout(
    title = paste0("Correlation Matrix — ", corr_window,
                   "-day window ending ", format(corr_snap)),
    xaxis = list(title = ""),
    yaxis = list(title = "", autorange = "reversed")
  )
```

### 5.2 Rolling Pairwise Correlation

The rolling correlation between two markets reveals structural changes:
e.g. WTI and Brent were nearly perfectly correlated before the US lifted
its crude export ban in 2015; the relationship shifted afterwards.

```{r}
market_a <- "CL"
market_b <- "BRN"

pair_data <- wide_ret |>
  select(date, all_of(c(market_a, market_b))) |>
  rename(col_a = all_of(market_a), col_b = all_of(market_b)) |>
  filter(!is.na(col_a), !is.na(col_b)) |>
  mutate(
    roll_corr = slide2_dbl(
      col_a, col_b,
      function(.x, .y) {
        if (length(.x) < 10L) return(NA_real_)
        cor(.x, .y, use = "complete.obs")
      },
      .before   = corr_window - 1L,
      .complete = TRUE
    )
  ) |>
  filter(!is.na(roll_corr))

plot_ly(data = pair_data, x = ~date, y = ~roll_corr,
        type = "scatter", mode = "lines",
        line = list(color = "#2980b9"),
        hovertemplate = "%{x|%Y-%m-%d}<br>Corr: %{y:.2f}<extra></extra>") |>
  layout(
    title  = paste0(market_a, " vs ", market_b,
                    " — ", corr_window, "-day rolling correlation"),
    xaxis  = list(title = "Date"),
    yaxis  = list(title = "Correlation", range = c(-1, 1))
  )
```

### 5.3 PCA — Principal Component Analysis of Market Co-Movement

PCA reduces the correlation structure of all selected markets into a small
number of uncorrelated **principal components (PCs)**.

- **PC1** typically captures the broad "risk-on / risk-off" factor — when
  crude, distillates, and gasoline all move together (high-vol regimes, demand
  shocks).
- **PC2** often separates natural gas (weather-driven, regional) from the oil
  complex — it tells you when NG is diverging from crude.
- **Loadings** show each market's contribution to each PC.  A large positive
  loading means the market moves in the same direction as the PC; a near-zero
  loading means it is largely orthogonal to that factor.
- **Scree plot**: the % variance explained by each PC.  If PC1 explains 70%+,
  markets are highly co-integrated.  If variance is spread across 3–4 PCs,
  markets have more idiosyncratic drivers.
- **PC1 score time-series**: a single number per day representing the
  "market stress" factor.  Spikes correspond to COVID March 2020, the 2022
  Russia-Ukraine supply shock, etc.

```{r}
# Build the same wide returns matrix used by the app
pca_markets <- c("CL", "BRN", "NG", "HO", "RB")

wide_pca <- futures |>
  filter(market %in% pca_markets, contract == 1) |>
  arrange(market, date) |>
  group_by(market) |>
  mutate(log_ret = c(NA_real_, diff(log(value)))) |>
  ungroup() |>
  filter(!is.na(log_ret)) |>
  pivot_wider(id_cols = date, names_from = market, values_from = log_ret) |>
  arrange(date)

# Drop rows with any NA (prcomp requires complete cases)
mat_full  <- as.matrix(wide_pca[, -1])
ok_rows   <- complete.cases(mat_full)
mat_clean <- mat_full[ok_rows, ]
dates_pca <- wide_pca$date[ok_rows]

# Run PCA (center + scale so all markets are unit-variance)
pca <- prcomp(mat_clean, center = TRUE, scale. = TRUE)

# How many PCs to show
n_pcs <- ncol(pca$rotation)
```

**Scree plot** — variance explained per PC and cumulative:

```{r}
var_exp <- pca$sdev^2 / sum(pca$sdev^2) * 100
cum_var <- cumsum(var_exp)
pc_labels <- paste0("PC", seq_len(n_pcs))

plot_ly() |>
  add_bars(
    x = pc_labels, y = var_exp,
    name   = "Explained (%)",
    marker = list(color = "#2980b9"),
    hovertemplate = "%{x}: %{y:.1f}%<extra></extra>"
  ) |>
  add_trace(
    x = pc_labels, y = cum_var,
    type = "scatter", mode = "lines+markers",
    name  = "Cumulative (%)",
    yaxis = "y2",
    line  = list(color = "#e74c3c"),
    hovertemplate = "%{x}: %{y:.1f}%<extra></extra>"
  ) |>
  layout(
    title  = "Scree Plot — Variance Explained by Each PC",
    xaxis  = list(title = ""),
    yaxis  = list(title = "Variance explained (%)"),
    yaxis2 = list(title = "Cumulative (%)", overlaying = "y",
                  side = "right", range = c(0, 100)),
    legend = list(orientation = "h", x = 0, y = -0.2),
    bargap = 0.3
  )
```

**Loadings** — which markets drive each PC:

```{r}
rot <- pca$rotation[, 1:min(3L, n_pcs), drop = FALSE]
mkts <- rownames(rot)

plot_ly() |>
  add_bars(x = mkts, y = rot[, 1], name = "PC1",
           marker = list(color = "#2980b9"),
           hovertemplate = "%{x}: %{y:.3f}<extra></extra>") |>
  add_bars(x = mkts, y = rot[, 2], name = "PC2",
           marker = list(color = "#e67e22"),
           hovertemplate = "%{x}: %{y:.3f}<extra></extra>") |>
  add_bars(x = mkts, y = rot[, 3], name = "PC3",
           marker = list(color = "#27ae60"),
           hovertemplate = "%{x}: %{y:.3f}<extra></extra>") |>
  layout(
    title   = "PC Loadings — first 3 principal components",
    barmode = "group",
    xaxis   = list(title = ""),
    yaxis   = list(title = "Loading"),
    legend  = list(orientation = "h", x = 0, y = -0.25)
  )
```

**PC1 score time-series** — the dominant co-movement factor over time:

```{r}
pc1_scores <- pca$x[, 1]

plot_ly(x = dates_pca, y = pc1_scores,
        type = "scatter", mode = "lines",
        line = list(color = "#8e44ad"),
        hovertemplate = "%{x|%Y-%m-%d}<br>PC1 score: %{y:.3f}<extra></extra>") |>
  layout(
    title = "PC1 Score — daily time series (dominant co-movement factor)",
    xaxis = list(title = "Date"),
    yaxis = list(title = "PC1 score")
  )
```

---

## 6. Seasonality

Energy markets have well-known seasonal patterns driven by demand cycles:
natural gas spikes in winter (heating) and summer (cooling); heating oil
draws in autumn as inventories are built before winter; refined products
tend to rally ahead of the summer driving season.

### 6.1 Average Monthly Log Return

```{r}
month_labels <- c("Jan","Feb","Mar","Apr","May","Jun",
                  "Jul","Aug","Sep","Oct","Nov","Dec")

seasonal_markets <- c("CL", "NG", "HO", "RB")

monthly_avg <- futures |>
  filter(market %in% seasonal_markets, contract == 1) |>
  arrange(market, date) |>
  group_by(market) |>
  mutate(log_ret = c(NA_real_, diff(log(value)))) |>
  ungroup() |>
  filter(!is.na(log_ret)) |>
  mutate(month = month(date)) |>
  group_by(market, month) |>
  summarise(avg_ret = mean(log_ret, na.rm = TRUE), .groups = "drop") |>
  mutate(month_lbl = month_labels[month])

p_seas <- plot_ly()
for (mkt in seasonal_markets) {
  sub_df <- filter(monthly_avg, market == mkt) |> arrange(month)
  p_seas <- add_trace(
    p_seas,
    data          = sub_df,
    x             = ~month_lbl,
    y             = ~avg_ret,
    type          = "bar",
    name          = mkt,
    hovertemplate = "%{x}: %{y:.2%}<extra></extra>"
  )
}
p_seas |>
  layout(
    barmode   = "group",
    title     = "Average Monthly Log Return — Front Month (2007–present)",
    xaxis     = list(title = "Month", categoryorder = "array",
                     categoryarray = month_labels),
    yaxis     = list(title = "Avg. Log Return", tickformat = ".1%"),
    legend    = list(orientation = "h", y = -0.25)
  )
```

### 6.2 Monthly Return Distribution (Box Plot)

Box plots show not just the average but the full distribution — important
for communicating tail risk to a risk management audience.

```{r}
monthly_rets <- futures |>
  filter(market %in% seasonal_markets, contract == 1) |>
  arrange(market, date) |>
  group_by(market) |>
  mutate(log_ret = c(NA_real_, diff(log(value)))) |>
  ungroup() |>
  filter(!is.na(log_ret)) |>
  mutate(month_lbl = month_labels[month(date)])

p_box <- plot_ly()
for (mkt in seasonal_markets) {
  sub_df <- filter(monthly_rets, market == mkt)
  p_box <- add_trace(
    p_box,
    data = sub_df,
    x    = ~month_lbl,
    y    = ~log_ret,
    type = "box",
    name = mkt,
    hovertemplate = "%{x}<br>%{y:.2%}<extra></extra>"
  )
}
p_box |>
  layout(
    title  = "Monthly Return Distributions — Front Month",
    xaxis  = list(title = "Month", categoryorder = "array",
                  categoryarray = month_labels),
    yaxis  = list(title = "Log Return", tickformat = ".1%"),
    legend = list(orientation = "h", y = -0.25)
  )
```

---

## 7. Hedge Ratios

A hedge ratio answers: "for every unit of exposure in market A, how many
units of market B do I need to offset the risk?"  We estimate it as the
slope (beta) from a rolling OLS regression of market A prices on market B
prices.  Rolling estimation captures how the relationship changes over time.

### 7.1 Cross-Market Hedge Ratio — CL vs BRN

WTI and Brent are close substitutes; a beta near 1 means one barrel of
WTI hedges roughly one barrel of Brent.  Deviations signal spread regime
changes — e.g. WTI traded at a deep discount during the 2011 Cushing glut.

```{r}
hedge_window <- 63  # ~1 quarter

# Helper: rolling OLS beta using slide2_dbl
rolling_beta <- function(y_vec, x_vec, window) {
  slide2_dbl(
    y_vec, x_vec,
    function(.y, .x) {
      sub <- data.frame(y = .y, x = .x)
      sub <- sub[complete.cases(sub), ]
      if (nrow(sub) < 10L) return(NA_real_)
      coef(lm(y ~ x, data = sub))[["x"]]
    },
    .before   = window - 1L,
    .complete = TRUE
  )
}

# Front-month prices in wide format
front_wide <- futures |>
  filter(contract == 1) |>
  select(date, market, value) |>
  pivot_wider(names_from = market, values_from = value) |>
  arrange(date)

cross_beta <- front_wide |>
  filter(!is.na(CL), !is.na(BRN)) |>
  mutate(beta = rolling_beta(CL, BRN, hedge_window)) |>
  filter(!is.na(beta))

plot_ly(data = cross_beta, x = ~date, y = ~beta,
        type = "scatter", mode = "lines",
        line = list(color = "#e74c3c"),
        hovertemplate = "%{x|%Y-%m-%d}<br>Beta: %{y:.3f}<extra></extra>") |>
  add_trace(x = range(cross_beta$date), y = c(1, 1),
            type = "scatter", mode = "lines",
            line = list(color = "grey", dash = "dot"),
            showlegend = FALSE) |>
  layout(
    title = paste0("CL ~ BRN hedge ratio — ", hedge_window, "-day rolling OLS"),
    xaxis = list(title = "Date"),
    yaxis = list(title = "Beta (hedge ratio)")
  )
```

### 7.2 Term-Structure Hedge Ratio — CL Contract 1 vs Contracts 2–6

A market maker hedging a front-month position with deferred contracts
needs to know how the hedge ratio changes along the curve.  A beta < 1
on C1 vs C3 means the front month moves more than the third contract —
typical in backwardation during supply shocks.

```{r}
ts_market   <- "CL"
max_contract <- 6

c1_data <- futures |>
  filter(market == ts_market, contract == 1) |>
  select(date, c1_price = value)

p_ts <- plot_ly()

for (cn in seq(2L, max_contract)) {
  cn_data <- futures |>
    filter(market == ts_market, contract == cn) |>
    select(date, cn_price = value) |>
    inner_join(c1_data, by = "date") |>
    arrange(date) |>
    mutate(beta = rolling_beta(c1_price, cn_price, hedge_window)) |>
    filter(!is.na(beta))

  if (nrow(cn_data) == 0L) next

  p_ts <- add_trace(
    p_ts,
    data          = cn_data,
    x             = ~date,
    y             = ~beta,
    type          = "scatter",
    mode          = "lines",
    name          = paste0("C1 vs C", cn),
    hovertemplate = paste0("C1 vs C", cn,
                           "<br>%{x|%Y-%m-%d}<br>Beta: %{y:.3f}<extra></extra>")
  )
}

p_ts |>
  layout(
    title  = paste0(ts_market, " — term-structure hedge ratios (", hedge_window, "-day)"),
    xaxis  = list(title = "Date"),
    yaxis  = list(title = "Beta"),
    legend = list(orientation = "h", y = -0.25)
  )
```

---

## 8. Cushing Storage vs WTI Calendar Spread

Cushing storage utilization is the fundamental driver of WTI curve shape.
When tanks are nearly full, holders pay a premium to store barrels in deferred
contracts → **contango** (negative C1-C2 spread).
When storage is tight, nearby barrels command a scarcity premium → **backwardation**
(positive C1-C2 spread).

`RTL::cushing$storage` gives us weekly EIA storage data from 2011 with
utilization (stocks / capacity) pre-calculated.

```{r}
storage <- RTL::cushing$storage |>
  select(date, utilization, c1c2) |>
  arrange(date)

cat("Storage date range:", format(min(storage$date)), "→", format(max(storage$date)), "\n")
cat("Rows:", nrow(storage), "\n")
head(storage)
```

```{r}
# Dual-axis chart: utilization (left) vs C1-C2 spread (right)
plot_ly() |>
  add_trace(
    data          = storage,
    x             = ~date,
    y             = ~utilization,
    type          = "scatter",
    mode          = "lines",
    name          = "Cushing utilization",
    fill          = "tozeroy",
    fillcolor      = "rgba(41,128,185,0.12)",
    line          = list(color = "#2980b9", width = 1.5),
    hovertemplate = "%{x|%Y-%m-%d}<br>Utilization: %{y:.1%}<extra></extra>"
  ) |>
  add_trace(
    data          = storage,
    x             = ~date,
    y             = ~c1c2,
    type          = "scatter",
    mode          = "lines",
    name          = "C1–C2 spread ($/bbl)",
    yaxis         = "y2",
    line          = list(color = "#e74c3c", width = 1.5),
    hovertemplate = "%{x|%Y-%m-%d}<br>C1–C2: $%{y:.2f}<extra></extra>"
  ) |>
  add_trace(
    x = range(storage$date), y = c(0, 0),
    type = "scatter", mode = "lines",
    yaxis = "y2",
    line = list(color = "grey", dash = "dot", width = 1),
    showlegend = FALSE, hoverinfo = "skip"
  ) |>
  layout(
    title  = "Cushing Storage Utilization vs WTI Calendar Spread (C1–C2)",
    xaxis  = list(title = "Date"),
    yaxis  = list(title = "Storage Utilization", tickformat = ".0%",
                  range = c(0.4, 1.0)),
    yaxis2 = list(title = "C1–C2 Spread ($/bbl)", overlaying = "y",
                  side = "right", showgrid = FALSE, zeroline = FALSE),
    legend = list(orientation = "h", y = -0.2),
    hovermode = "x unified"
  )
```

---

## 9. EIA Inventory Stocks — All Markets

`RTL::eiaStocks` bundles four weekly EIA inventory series that map directly
to our markets:

| Series | Market | Unit |
|---|---|---|
| `CrudeCushing` | CL / HTT | Thousand barrels |
| `NGLower48` | NG | Bcf |
| `ULSD` | HO | Thousand barrels |
| `Gasoline` | RB | Thousand barrels |

For a Risk Manager, the key question is not the absolute stock level but
**how it compares to the seasonal norm** — the 5-year average for the same
week of the year. A deficit going into peak demand season drives prices up;
a surplus caps them.

### 9.1 Data Overview

```{r}
stocks <- RTL::eiaStocks |>
  dplyr::arrange(series, date)

# Date range, row count, and latest value per series
stocks |>
  dplyr::group_by(series) |>
  dplyr::summarise(
    from   = min(date),
    to     = max(date),
    rows   = dplyr::n(),
    latest = dplyr::last(value),
    .groups = "drop"
  )
```

### 9.2 Stocks vs 5-Year Seasonal Average

The 5-year average is computed per calendar week using only the prior 5 years
(no look-ahead), matching the EIA's own published methodology.
The surplus/deficit column shows how many units above or below the seasonal
norm inventories sit today.

```{r}
stocks_norm <- stocks |>
  dplyr::mutate(
    week = lubridate::week(date),
    year = lubridate::year(date)
  ) |>
  dplyr::group_by(series, week) |>
  dplyr::mutate(
    five_yr_avg     = slider::slide_index_dbl(
      value, year,
      ~ mean(.x, na.rm = TRUE),
      .before = 5L, .after = -1L
    ),
    surplus_deficit = value - five_yr_avg
  ) |>
  dplyr::ungroup()

# Spot check latest reading per series
stocks_norm |>
  dplyr::group_by(series) |>
  dplyr::slice_tail(n = 1) |>
  dplyr::select(series, date, value, five_yr_avg, surplus_deficit)
```

### 9.3 NG Storage vs 5-Year Average

Natural Gas is the most storage-sensitive market in the complex. The EIA
reports working gas in storage every Thursday. A deficit vs the 5-year
average heading into the November–March heating season is the primary
fundamental driver of NG price rallies.

```{r}
# Drop rows where the 5-yr average hasn't warmed up yet (first ~5 years)
ng_stocks <- dplyr::filter(stocks_norm, series == "NGLower48",
                            !is.na(five_yr_avg))

plotly::plot_ly() |>
  plotly::add_trace(
    data = ng_stocks, x = ~date, y = ~five_yr_avg,
    type = "scatter", mode = "lines", name = "5-yr avg",
    line = list(color = "grey", dash = "dash", width = 1.5),
    hovertemplate = "%{x|%Y-%m-%d}<br>5-yr avg: %{y:.0f} Bcf<extra></extra>"
  ) |>
  plotly::add_trace(
    data = ng_stocks, x = ~date, y = ~value,
    type = "scatter", mode = "lines", name = "Actual",
    line = list(color = "#e74c3c", width = 2),
    hovertemplate = "%{x|%Y-%m-%d}<br>Actual: %{y:.0f} Bcf<extra></extra>"
  ) |>
  plotly::add_trace(
    data = ng_stocks, x = ~date, y = ~surplus_deficit,
    type = "bar", name = "Surplus / Deficit vs 5-yr avg",
    yaxis = "y2",
    marker = list(
      color = dplyr::case_when(
        ng_stocks$surplus_deficit >= 0 ~ "rgba(39,174,96,0.5)",
        TRUE                           ~ "rgba(192,57,43,0.5)"
      )
    ),
    hovertemplate = "%{x|%Y-%m-%d}<br>+/- 5yr: %{y:.0f} Bcf<extra></extra>"
  ) |>
  plotly::layout(
    title     = "NG Lower 48 Working Gas in Storage vs 5-Year Average",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Working Gas (Bcf)"),
    yaxis2    = list(title = "Surplus / Deficit (Bcf)", overlaying = "y",
                     side = "right", showgrid = FALSE, zeroline = TRUE),
    hovermode = "x unified",
    legend    = list(orientation = "h", y = -0.2)
  )
```

### 9.4 Petroleum Products — ULSD and Gasoline vs 5-Year Average

```{r}
prod_stocks <- dplyr::filter(stocks_norm, series %in% c("ULSD", "Gasoline"),
                              !is.na(five_yr_avg))

plotly::plot_ly() |>
  plotly::add_trace(
    data = dplyr::filter(prod_stocks, series == "ULSD"),
    x = ~date, y = ~surplus_deficit,
    type = "scatter", mode = "lines",
    name = "ULSD surplus/deficit (HO proxy)",
    line = list(color = "#e67e22", width = 1.5),
    hovertemplate = "ULSD<br>%{x|%Y-%m-%d}<br>+/- 5yr: %{y:.0f} Mbbl<extra></extra>"
  ) |>
  plotly::add_trace(
    data = dplyr::filter(prod_stocks, series == "Gasoline"),
    x = ~date, y = ~surplus_deficit,
    type = "scatter", mode = "lines",
    name = "Gasoline surplus/deficit (RB proxy)",
    line = list(color = "#8e44ad", width = 1.5),
    hovertemplate = "Gasoline<br>%{x|%Y-%m-%d}<br>+/- 5yr: %{y:.0f} Mbbl<extra></extra>"
  ) |>
  plotly::add_trace(
    x = range(prod_stocks$date), y = c(0, 0),
    type = "scatter", mode = "lines",
    line = list(color = "grey", dash = "dot", width = 1),
    showlegend = FALSE, hoverinfo = "skip"
  ) |>
  plotly::layout(
    title     = "US Product Stocks vs 5-Year Seasonal Average (Thousand Barrels)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Surplus / Deficit vs 5-yr avg (Mbbl)"),
    hovermode = "x unified",
    legend    = list(orientation = "h", y = -0.2)
  )
```

---

## 10. EIA API vs RTL::eiaStocks Comparison

### 10.1 Find matching tickers in RTL::tickers_eia

Search for storage-related series to identify the correct ticker strings
for Cushing crude, NG Lower 48, distillate/ULSD, and gasoline.

```{r}
# Show all unique categories to understand the structure
unique(RTL::tickers_eia$category)
```

```{r}
# Filter for storage/stocks series
RTL::tickers_eia |>
  dplyr::filter(grepl("stock|storage|cushing|inventory",
                       tick.r, ignore.case = TRUE)) |>
  dplyr::select(tick.eia, tick.r, product, category, sd_category) |>
  print(n = 50)
```

### 10.2 Fetch via EIA API

Replace the ticker strings below with the ones identified above, and paste
your EIA API key where indicated.

```{r}
EIA_KEY <- "PASTE_YOUR_KEY_HERE"

api_data <- RTL::eia2tidy_all(
  tickers = tibble::tribble(
    ~ticker,                         ~name,
    "PET.W_EPC0_SAX_YCUOK_MBBL.W", "CrudeCushing",
    "NG.NW2_EPG0_SWO_R48_BCF.W",   "NGLower48",
    "PET.WDISTUS1.W",               "Distillate",
    "PET.WGTSTUS1.W",               "Gasoline"
  ),
  key  = EIA_KEY,
  long = TRUE
)

api_data |>
  dplyr::group_by(series) |>
  dplyr::summarise(from = min(date), to = max(date),
                   rows = dplyr::n(), latest = dplyr::last(value))
```

### 10.3 Compare with RTL::eiaStocks

```{r}
RTL::eiaStocks |>
  dplyr::group_by(series) |>
  dplyr::summarise(from = min(date), to = max(date),
                   rows = dplyr::n(), latest = dplyr::last(value))
```

### 10.4 Side-by-side value check (after filling tickers above)

```{r}
# Map: api series name -> RTL::eiaStocks series name
series_map <- tibble::tribble(
  ~api_name,      ~rtl_name,
  "CrudeCushing", "CrudeCushing",
  "NGLower48",    "NGLower48",
  "Distillate",   "ULSD",
  "Gasoline",     "Gasoline"
)

purrr::pmap_dfr(series_map, function(api_name, rtl_name) {
  api_s <- dplyr::filter(api_data, series == api_name)
  rtl_s <- dplyr::filter(RTL::eiaStocks, series == rtl_name)

  dplyr::inner_join(api_s, dplyr::select(rtl_s, date, rtl_value = value),
                    by = "date") |>
    dplyr::mutate(diff = value - rtl_value) |>
    dplyr::summarise(
      api_series   = api_name,
      rtl_series   = rtl_name,
      n_matched    = dplyr::n(),
      api_from     = min(date),
      api_to       = max(date),
      rtl_rows     = nrow(rtl_s),
      max_abs_diff = max(abs(diff), na.rm = TRUE),
      mean_diff    = mean(diff, na.rm = TRUE)
    )
})
```

---

## Summary

| Tab | Key question |
|---|---|
| Forward Curve | Contango or backwardation? How has the term structure shifted? |
| Volatility | Where is vol elevated — near-term or far-term? Which crisis shows up? |
| Co-Dynamics — Correlation | Which markets move together? When did that relationship break? |
| Co-Dynamics — PCA | How many independent risk factors drive these markets? What is each factor? |
| Seasonality | Which months are systematically bullish or bearish? |
| Hedge Ratios | How many units of the hedge instrument offset one unit of exposure? |
